Clinical prediction with localized modeling using similarity-based cohorts: A scoping review

### 2023-11-25 docs/Localized-Modeling/Localized-Modeling.Rmd
relocate bibliography files + begin submission to air

---
title: "Clinical prediction with localized modeling using similarity-based cohorts: A scoping review"
author:
  - name: Adi Cohen
    email: ac4569@my.nsu.edu
    affiliation: 1
  - name: Patti McCall-Junkin
    email: pattimccall.junkin@gmail.com
    affiliation: 2
  - name: Jason Cory Brunson
    email: jason.brunson@medicine.ufl.edu
    affiliation: 3
    correspondingauthor: true
address:
  - code: 1
    address: College of Osteopathic Medicine, Nova Southeastern University, 3200 S University Drive, Davie, FL 33328
  - code: 2
    address: Smathers Libraries Academic Research & Consulting Services, University at Florida, P. O. Box 117000, Gainesville, FL 32611-7000
  - code: 3
    address: Laboratory for Systems Medicine, University of Florida, P.O. Box 100225 JHMHC, Gainesville, FL 32610-0225
abstract: |
  Background:
  Applications of classical case-based reasoning (CBR) to clinical and health tasks have given rise to a family of techniques in which a statistical model is fitted to a subset of labeled cases, selected for relevance or similarity to an unlabeled case to which the model is applied. These localized models, as we propose to call them, may perform as well as "black-box" models while retaining direct interpretations. Their adoption has been hindered by fragmented development and implementation challenges. Our goal in this scoping review was to describe the clinical and health applications of localized models to date and to derive a general framework for their design and evaluation.

  Methods:
  We searched four electronic bibliographic platforms (PubMed, Web of Science, Academic Search Premier, and MathSciNet) during 2021 July 19–22. We screened and reviewed entries based on four criteria intended to specify applications of localized models to tasks involving analysis of clinical or health data. Two authors divided title/abstract screening and collaboratively reviewed screened entries for inclusion. We also screened and reviewed a seed set that inspired the review and traced methodological citations from included entries. We discussed and summarized settings, tasks, and tools; identified and tabulated themes; and synthesized the methods into a general framework. We received no specific funding for this work.

  Results:
  Of 1,309 search results, 328 were reviewed. Of these, combined with 43 publications that seeded the review and 1 obtained by citation tracking, 25 were included in the review and methodological synthesis. The specificity of several search terms was poor, and inter-rater reliability was low. Entries were published from 1997 to 2021, with half of these since 2015. The most common tasks were prognosis and diagnosis. The studies overwhelmingly focused on clinical data, occasionally including laboratory and image-derived data. Only three studies applied localized models to tasks other than prediction. Several studies commented on memory and runtime costs, and four proposed partial solutions. While few methods were reused and no results were reproduced, a general technique that specializes to most of those encountered was straightforward to describe.

  Conclusions:
  Localized models have potential to improve the performance of models used in clinical decision support tools while maintaining interpretability, provided computational hurdles can be overcome. Our search undoubtedly failed to capture all studies that used methods of this kind, but those we found follow a consistent enough design to be reproduced by a general-purpose tool. We hope that our review and synthesis will spur future work on efficiency, reproducibility, and user interactions and needs.
keywords: 
  - case-based reasoning
  - localized modeling
  - nearest neighbors
  - clinical decision support
  - scoping review
journal: "Artificial Intelligence In Medicine"
date: "`r format(Sys.Date(), format = '%Y %b %d')`"
classoption: preprint, 3p, authoryear
bibliography:
  - SR-bibliography.bib
  - SR-0-Found.bib
  - SR-2-Included.bib
  - SR-3-Tracked.bib
linenumbers: true
numbersections: true
# Use a CSL with `citation_package = "default"`
# csl: https://www.zotero.org/styles/elsevier-harvard
# csl: https://www.zotero.org/styles/artificial-intelligence-in-medicine
csl: artificial-intelligence-in-medicine.csl
output: 
  rticles::elsevier_article:
    keep_tex: true
    latex_engine: xelatex
    # citation_package: natbib
    citation_package: "default"
header-includes:
  \usepackage{lscape}
  \newcommand{\landscapebegin}{\begin{landscape}}
  \newcommand{\landscapeend}{\end{landscape}}

---

@@

@@

Clinical decision support (CDS) systems have been a fixture of artificial intelligence (AI) research since inception. As machine learning (ML) emerged, clinical modeling tasks became some of the most common and celebrated applications. A technique core to clinical applications of both AI and ML is _similarity-matching_, which relies on a _patient similarity measure_[^psm] to extract a cohort of past or training cases used to inform the diagnosis, prognosis, or care of a new or test case. Similarity-matching powers the retrieval step of most case-based reasoning (CBR) systems and provides the neighborhoods from which nearest neighbor (NN) models derive predictions. Whereas NN predictions are generated automatically via averaging (regression) or voting (classification), traditional CBR tools return the retrieved cases to inform human decision-making.

@@

We discuss in this review a hybrid computational approach we call _localized modeling_ (see Section \ref{coherence}) that appears to have been introduced independently several times. The approach fits ML models, including regularized families of generalized regression models (GLMs), to cohorts retrieved via similarity matching. Whereas CBR and NN prediction use patient similarity to obtain an ideally homogeneous subpopulation matched to the index case, in localized modeling one seeks a relevant yet heterogeneous subpopulation, each member of which may deviate from the index case.
A prediction or inference is then obtained from a model fitted to this _similarity cohort_.
The use of similarity cohorts distinguishes localized modeling from classical GLMs that still dominate analysis in clinical trials and comparative effectiveness research, as well as from many cutting-edge ML models that are fitted to whole populations, while the cohorts' heterogeneity and its utility to the model distinguish the approach from CBR.
In principle, this allows interpretable models like GLMs to identify and measure the effects of determining factors that are specific to the index case.
In the reviewed studies, localized modeling shows some promise toward simultaneously achieving the performance improvements of ML and the interpretability of classical models.

<!--
In multivariate settings, because similarity decreases with difference along each factor, included cases that differ more with respect to some factors will tend to differ less with respect to others. Consequently, (1) the distribution of a factor in the subpopulation will tend to be centered near its value at the index case, and (2) the joint distribution of a pair of factors will tend to be sparse away from both center axes, i.e. when both factors differ from their values at the index case. This should produce localized models that are less responsive to deviation along multiple uncorrelated factors, so that the influence of each factor in the model reflects more its independent effect than its interactions with other factors.
-->

One memorable aspiration of clinical informatics was the "green button" that would, in response to a query, automatically retrieve patient data from an institution’s records with which to conduct on-demand retrospective studies for an individual patient @Longhurst2014. Localized modeling has recently been put forth as one answer to this appeal. Studies using localized models have been limited by computational cost—a naïve implementation involves fitting, optimizing, and evaluating a model on a separate population for each training case—and recent studies have described algorithms and implementations designed to reduce this cost. As greater access to health data, informatical knowledge, algorithmic efficiency, and computational power continue to spur innovation in CDS, we may expect localized modeling to become more accessible and efficient, its advantages better understood, and its disadvantages ameliorated.

The motivation for this scoping review is to describe the settings in and problems with which localized models have been tasked to date, and to synthesize existing methods into a general framework. Along the way, we attempt to reconcile terminology, summarize motivations and evaluations, and propose valuable follow-up work.

@@

```{r framework, fig.align='center', out.width="80%", echo=FALSE}
#| fig.cap =
#| "General framework for localized models."
knitr::include_graphics("../../fig/fig-typology.pdf")
```

@@

# Acknowledgments

>>>>

---
# Supported options: 
#   sn-nature:       Style for submissions to Nature Portfolio journals
#   sn-basic:        Basic Springer Nature Reference Style/Chemistry Reference Style
#   sn-mathphys:     Math and Physical Sciences Reference Style
#   sn-aps:          American Physical Society (APS) Reference Style
#   sn-vancouver:    Vancouver Reference Style
#   sn-apa:          APA Reference Style 
#   sn-chicago:      Chicago-based Humanities Reference Style
#   default:         Default

classoptions: 
  - sn-nature      
  # - Numbered      # Optional: Use numbered references instead of namedate references (only for sn-nature, sn-basic, sn-vancouver, sn-chicago, sn-mathphys or sn-nature)
  # - referee       # Optional: Use double line spacing 
  # - lineno        # Optional: Add line numbers
  # - iicol         # Optional: Double column layour

title: "Clinical prediction with localized modeling using similarity-based cohorts: A scoping review"
titlerunning: Clinical Prediction with Localized Modeling
authors: 
  - firstname: Adi
    lastname: Cohen
    degrees: BS?
    email: ac4569@my.nsu.edu
    affiliation: 1
  - firstname: Patti
    lastname: McCall-Junkin
    degrees: MA, MLS
    email: pattimccall.junkin@gmail.com
    affiliation: 2
  - firstname: Jason Cory
    lastname: Brunson
    degrees: PhD
    email: jason.brunson@medicine.ufl.edu
    orcid: 0000-0003-3126-9494
    affiliation: 3
    corresponding: TRUE
affiliations:
  - number: 1
    info:
      orgdiv: College of Osteopathic Medicine
      orgname: Nova Southeastern University
    address:
        street: 3200 S University Drive
        city: Davie
        postcode: 33328
        state: FL
        country: USA
  - number: 2
    info:
      orgdiv: Smathers Libraries Academic Research & Consulting Services
      orgname: University of Florida
    address:
        street: P.O. Box 117000
        city: Gainesville
        postcode: 32611-7000
        state: FL
        country: USA
  - number: 3
    corresponding: TRUE
    info:
      orgdiv: Laboratory for Systems Medicine
      orgname: University of Florida
    address:
        street: P.O. Box 100225 JHMHC
        city: Gainesville
        postcode: 32610-0225
        state: FL
        country: USA
keywords:
  - case-based reasoning
  - localized modeling
  - nearest neighbors
  - clinical decision support
  - scoping review
# pacs: 
#   jel:
#     - "D8"
#     - "H51"
#   msc:
#     - "35A01"
#     - "65L10"

# Sample for structured abstract
abstract: |
  Background:
  Applications of classical case-based reasoning (CBR) have given rise to a family of techniques we call “localized models”, in which a statistical model is fitted to a neighborhood of similar labeled cases. We aim to describe clinical and health applications of localized models to date and propose a general framework for their design and evaluation.

  Methods:
  We searched four bibliographic platforms during 2021 July 19–22. We set four eligibility criteria to identify applications of localized models to clinical and health tasks. Two authors divided title/abstract screening and reviewed screened entries for inclusion. We discussed settings, tasks, and tools; identified and tabulated themes; and synthesized the methods into a general framework.

  Results:
  Of 1,309 search results, 328 were reviewed, then combined with 43 publications that seeded the review and 1 obtained by citation tracking. 25 were included, published 1997–2021. The specificity of search terms was poor, and inter-rater reliability was low. Almost all models were predictive, the most common tasks being prognosis and diagnosis. Most studies used clinical, occasionally laboratory and image, data. Several addressed memory and runtime costs. A general technique that specializes to most of those reviewed was straightforward to describe.

  Conclusions:
  Localized models have potential to improve the performance of clinical decision support tools while maintaining interpretability, provided computational hurdles are overcome, and they follow a consistent pattern that could be implemented in a general-purpose way. We hope that our review will spur future work on efficiency, reproducibility, and user needs.

bibliography:
  - ../bibliography/SR-bibliography.bib
  - ../bibliography/SR-0-Found.bib
  - ../bibliography/SR-2-Included.bib
  - ../bibliography/SR-3-Tracked.bib

header-includes: |
  \usepackage{lscape}
  \newcommand{\landscapebegin}{\begin{landscape}}
  \newcommand{\landscapeend}{\end{landscape}}

  %% Per the spinger doc, new theorem styles can be included using built in style, 
  %% but it seems the don't work so commented below
  %\theoremstyle{thmstyleone}%
  \newtheorem{theorem}{Theorem}%  meant for continuous numbers
  %%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
  %% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
  \newtheorem{proposition}[theorem]{Proposition}%
  %%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.

  %% \theoremstyle{thmstyletwo}%
  \theoremstyle{remark}
  \newtheorem{example}{Example}%
  \newtheorem{remark}{Remark}%

  %% \theoremstyle{thmstylethree}%
  \theoremstyle{definition}
  \newtheorem{definition}{Definition}%

output:
  rticles::springer_article:
    keep_tex: true
    latex_engine: xelatex
    # citation_package: natbib
    citation_package: "default"
---

@@

### Acknowledgments {-}

@@

A main driver of the development of clinical information systems (CIS) and clinical decision support (CDS) tools has been to leverage the advantages of routinely collected health data toward improving care.
These advantages include their immediate accessibility through the institutional EHR, billing database, or other source; their specificity to the institution and the population it serves; and the closeness in time of the available data.
Though by definition not collected for research use, the secondary research use of routinely collected health data has been put forth, with the monicker "practice-based evidence", as an important complement to the paradigm of evidence-based practice.

Over the same period of time, interest in individualizing care from population-derived evidence-based recommendations to specific patient needs has driven the application of advanced computational tools, including artificial intelligence (AI). Whereas classical models produced formulae involving a limited set of data elements that would be uniformly applied to all patients, AI models often process hundreds or more variables in opaque ways to yield predictions that depend on not only their values but their associations and their interplay with respect to outcomes. The loss of straightforward interpretations of these models has limited their practical uptake and motivated the construction of explanatory statistics for opaque models as well as the development of more interpretable complex models.

One of few methodologies to address all three of these established needs is one we call _localized modeling_, which appears to have been introduced several times independently under different names, and in slightly different forms.
The approach is a specialized form of case-based reasoning (CBR) that relies on a patient similarity measure[^psm] to extract a cohort of past or training cases that then inform the diagnosis, prognosis, or care of a new or test case.
While traditional CBR returns these retrieved cases to inform human decision-making, the related nearest neighbors (NN) technique generates predictions from cohorts automatically via averaging (regression) or voting (classification) of their outcomes.
In contrast, localized modeling fits families of predictive models, most commonly generalized linear models (GLMs), to these retrieved similarity cohorts in order to generate predictions.
Localized models thus provide a way to fit fully interpretable models to large population data in a way that is personalized to each case.

@@

This approach harkens to the aspirational "green button" that would, in response to a query, automatically retrieve patient data from an institution’s records with which to conduct on-demand retrospective studies for an individual patient [@Longhurst2014].
Our motivations in this scoping review are to describe the settings in and problems with which this approach has been tasked, to attempt to organize them within a common framework, and to evaluate the promise they show toward achieving this or other needs in clinical informatics.
Along the way, we attempt to reconcile terminology, summarize motivations and evaluations, and propose valuable follow-up work.

@@

```{r framework, fig.align='center', out.width="80%", echo=FALSE}
#| fig.cap =
#| "General framework for localized models.
#|  Dashed arrows indicate optional steps."
knitr::include_graphics("../../fig/fig-typology.pdf")
```

@@

### 2023-11-25 docs/Manuscript/Manuscript.tex
relocate bibliography files + begin submission to air

  \title{Clinical prediction with localized modeling using
similarity-based cohorts: A scoping review}
    \author[1]{Adi Cohen%
  %
  }
   \ead{ac4569@my.nsu.edu} 
    \author[2]{Patti McCall-Junkin%
  %
  }
   \ead{pattimccall.junkin@gmail.com} 
    \author[3]{Jason Cory Brunson%
  \corref{cor1}%
  }
   \ead{jason.brunson@medicine.ufl.edu} 
      \affiliation[1]{College of Osteopathic Medicine, Nova Southeastern
University, 3200 S University Drive, Davie, FL 33328}
    \affiliation[2]{Smathers Libraries Academic Research \& Consulting
Services, University at Florida, P. O. Box 117000, Gainesville, FL
32611-7000}
    \affiliation[3]{Laboratory for Systems Medicine, University of
Florida, P.O. Box 100225 JHMHC, Gainesville, FL 32610-0225}
    \cortext[cor1]{Corresponding author}

@@

Clinical decision support (CDS) systems have been a fixture of
artificial intelligence (AI) research since inception. As machine
learning (ML) emerged, clinical modeling tasks became some of the most
common and celebrated applications. A technique core to clinical
applications of both AI and ML is \emph{similarity-matching}, which
relies on a \emph{patient similarity measure}\footnote{Similarity
  measures between more granular units of analysis, such as encounters
  and decision points, are often still referred to as patient similarity
  measures.} to extract a cohort of past or training cases used to
inform the diagnosis, prognosis, or care of a new or test case.
Similarity-matching powers the retrieval step of most case-based
reasoning (CBR) systems and provides the neighborhoods from which
nearest neighbor (NN) models derive predictions. Whereas NN predictions
are generated automatically via averaging (regression) or voting
(classification), traditional CBR tools return the retrieved cases to
inform human decision-making.

We discuss in this review a hybrid computational approach we call
\emph{localized modeling} (see Section \ref{coherence}) that appears to
have been introduced independently several times. The approach fits ML
models, including regularized families of generalized regression models
(GLMs), to cohorts retrieved via similarity matching. Whereas CBR and NN
prediction use patient similarity to obtain an ideally homogeneous
subpopulation matched to the index case, in localized modeling one seeks
a relevant yet heterogeneous subpopulation, each member of which may
deviate from the index case. A prediction or inference is then obtained
from a model fitted to this \emph{similarity cohort}. The use of
similarity cohorts distinguishes localized modeling from classical GLMs
that still dominate analysis in clinical trials and comparative
effectiveness research, as well as from many cutting-edge ML models that
are fitted to whole populations, while the cohorts' heterogeneity and
its utility to the model distinguish the approach from CBR. In
principle, this allows interpretable models like GLMs to identify and
measure the effects of determining factors that are specific to the
index case. In the reviewed studies, localized modeling shows some
promise toward simultaneously achieving the performance improvements of
ML and the interpretability of classical models.

One memorable aspiration of clinical informatics was the ``green
button'' that would, in response to a query, automatically retrieve
patient data from an institution's records with which to conduct
on-demand retrospective studies for an individual patient {[}1{]}.
Localized modeling has recently been put forth as one answer to this
appeal. Studies using localized models have been limited by
computational cost---a naïve implementation involves fitting,
optimizing, and evaluating a model on a separate population for each
training case---and recent studies have described algorithms and
implementations designed to reduce this cost. As greater access to
health data, informatical knowledge, algorithmic efficiency, and
computational power continue to spur innovation in CDS, we may expect
localized modeling to become more accessible and efficient, its
advantages better understood, and its disadvantages ameliorated.

The motivation for this scoping review is to describe the settings in
and problems with which localized models have been tasked to date, and
to synthesize existing methods into a general framework. Along the way,
we attempt to reconcile terminology, summarize motivations and
evaluations, and propose valuable follow-up work.

>>>>

  \title{Clinical prediction with localized modeling using
similarity-based cohorts: A scoping review}
    \author[1]{Adi Cohen%
  %
  }
   \ead{ac4569@my.nsu.edu} 
    \author[2]{Patti McCall-Junkin%
  %
  }
   \ead{pattimccall.junkin@gmail.com} 
    \author[3]{Jason Cory Brunson%
  \corref{cor1}%
  }
   \ead{jason.brunson@medicine.ufl.edu} 
      \affiliation[1]{College of Osteopathic Medicine, Nova Southeastern
University, 3200 S University Drive, Davie, FL 33328}
    \affiliation[2]{Smathers Libraries Academic Research \& Consulting
Services, University of Florida, P.O. Box 117000, Gainesville, FL
32611-7000}
    \affiliation[3]{Laboratory for Systems Medicine, University of
Florida, P.O. Box 100225 JHMHC, Gainesville, FL 32610-0225}
    \cortext[cor1]{Corresponding author}

@@

A main driver of the development of clinical information systems (CIS)
and clinical decision support (CDS) tools has been to leverage the
advantages of routinely collected health data toward improving care.
These advantages include their immediate accessibility through the
institutional EHR, billing database, or other source; their specificity
to the institution and the population it serves; and the closeness in
time of the available data. Though by definition not collected for
research use, the secondary research use of routinely collected health
data has been put forth, with the monicker ``practice-based evidence'',
as an important complement to the paradigm of evidence-based practice.

Over the same period of time, interest in individualizing care from
population-derived evidence-based recommendations to specific patient
needs has driven the application of advanced computational tools,
including artificial intelligence (AI). Whereas classical models
produced formulae involving a limited set of data elements that would be
uniformly applied to all patients, AI models often process hundreds or
more variables in opaque ways to yield predictions that depend on not
only their values but their associations and their interplay with
respect to outcomes. The loss of straightforward interpretations of
these models has limited their practical uptake and motivated the
construction of explanatory statistics for opaque models as well as the
development of more interpretable complex models.

One of few methodologies to address all three of these established needs
is one we call \emph{localized modeling}, which appears to have been
introduced several times independently under different names, and in
slightly different forms. The approach is a specialized form of
case-based reasoning (CBR) that relies on a patient similarity
measure\footnote{Similarity measures between more granular units of
  analysis, such as encounters and decision points, are often still
  referred to as patient similarity measures.} to extract a cohort of
past or training cases that then inform the diagnosis, prognosis, or
care of a new or test case. While traditional CBR returns these
retrieved cases to inform human decision-making, the related nearest
neighbors (NN) technique generates predictions from cohorts
automatically via averaging (regression) or voting (classification) of
their outcomes. In contrast, localized modeling fits families of
predictive models, most commonly generalized linear models (GLMs), to
these retrieved similarity cohorts in order to generate predictions.
Localized models thus provide a way to fit fully interpretable models to
large population data in a way that is personalized to each case.

This approach harkens to the aspirational ``green button'' that would,
in response to a query, automatically retrieve patient data from an
institution's records with which to conduct on-demand retrospective
studies for an individual patient {[}1{]}. Our motivations in this
scoping review are to describe the settings in and problems with which
this approach has been tasked, to attempt to organize them within a
common framework, and to evaluate the promise they show toward achieving
this or other needs in clinical informatics. Along the way, we attempt
to reconcile terminology, summarize motivations and evaluations, and
propose valuable follow-up work.
